---
title: "Net survival of rectal cancer patients"
author: "Author Name"
institute: "Author's affiliation"
date: "`r Sys.Date()`"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Patient cohort

Data frame `sire` includes observations of a simulated cohort of 8243 female Finnish rectal cancer patients diagnosed between 1993–2012.

## Statistical methods

The net survival was estimated by using the Pohar Perme estimator. Traditional direct age-standardisation was used for survival comparisons between diagnosis periods 1993–2002 and 2003–2012. The age standardisation was carried out using the International Cancer Survival Standard weights (ICSS1) in five age groups (0–44, 45–54, 55–64, 65–74 and 75 years and older).

## Estimation in R

```{r, echo=FALSE, results='hide', warning=FALSE}
library(Epi)
library(popEpi)
library(data.table)
library(knitr)

pm <- popEpi::popmort

sire$agegr <- cut(sire$dg_age,c(0,45,55,65,75,Inf),right=F)

Lex <- Lexis(entry = list(FUT = 0, AGE = dg_age, 
                          CAL = cal.yr(dg_date)),
             exit = list(CAL = cal.yr(ex_date)),
             data = sire,
             exit.status = factor(status, levels = 0:2,
                                  labels = c("alive", "canD", "othD")),
             merge = TRUE)

Lex$period <- cut( cal.yr(Lex$dg_date), c(1993,2003,2013), right=F,dig.lab = 4)
ICSS1 <- aggregate(ICSS1~cut(ICSS$age, 
                   c(0,45,55,65,75,Inf), right = FALSE), 
                   data = ICSS, FUN = sum)
w <- list(agegr = ICSS1$ICSS1)

BL <- list(FUT=seq(0, 15, by = 1/12))
names(pm) <- c("sex","CAL","AGE","haz")
```

Net survival was estimated by using `survtab` function in `popEpi` package. The estimation was done separately for the two subsets of data (calendar periods of diagnosis).

```{r, echo=TRUE, results='hide', warning=FALSE}
surv1993 <- survtab(
  Surv(time = FUT, event = lex.Xst%in%c("canD", "othD")) 
  ~ adjust(agegr), 
  data = Lex, subset=period=="[1993,2003)",
  pophaz = pm,surv.type = "surv.rel",
  relsurv.method = "pp",breaks = BL, weights=w)

surv2003 <- survtab(
  Surv(time = FUT, event = lex.Xst%in%c("canD", "othD")) 
  ~ adjust(agegr), 
  data = Lex, subset=period=="[2003,2013)",
  pophaz = pm,surv.type = "surv.rel",
  relsurv.method = "pp",breaks = BL, weights=w)
```

```{r, echo=FALSE, results='hide', warning=FALSE}
res1993 <- summary(surv1993,t=c(1,5,10,15))
res1993$period <- "1993-2002"
res2003 <- summary(surv2003,t=c(1,5,10))
res2003$period <- "2003-2012"

#combine estimates of the two periods
res <- rbind(res1993,res2003)

tb <- res[,.(period, Tstop, est=round(r.pp.as*100),ci=paste0("(",round(r.pp.as.lo*100),", ",round(r.pp.as.hi*100),")"))]

#from long to wide data format
tb <- dcast(tb,Tstop~period,value.var=c("est","ci"))

#reorder columns
tb <- tb[,c("Tstop","est_1993-2002","ci_1993-2002","est_2003-2012","ci_2003-2012")]
tb[is.na(tb)] <- "-"
```

## Estimates of net survival in table

Age-standardised net survival of rectal cancer patients at 1, 5, 10 and 15 years since diagnosis by calendar period of diagnosis

```{r, echo=FALSE}
kable(tb, col.names = c("Time (years)", "1993–2002", "95% CI", "2003–2012","95% CI"), align = c("l","r","r","r","r"))
```

## Net survival in figure

Figure shows the monthly estimates of cumulative age-standardised net survival of rectal cancer patients diagnosed in 1993–2002 and 2003–2012, respectively.

```{r, echo=FALSE}
par(mfrow=c(1,1),mar=c(4,4,1,1))
plot(0,0,type="n",xlim=c(0,15),ylim=c(0,100),las=1,xaxs="i",yaxs="i",
     ylab="Net survival (%)",
     xlab="Time since diagnosis (years)")
lines(c(0,surv1993[,Tstop]),c(100,surv1993[,r.pp.as*100]),col="red")
lines(c(0,surv2003[,Tstop]),c(100,surv2003[,r.pp.as*100]),col="blue")
legend("bottomright",c("2003–2012","1993–2002"),col=c("blue","red"),lty=1)
```

## Other output formats

In addition to this `html` document format, try to render this Quarto document to different presentation format by modifying the header:

-   html (`format: revealjs`)
-   powerpoint (`format: pptx`) 
-   pdf (`format: beamer`, N.B. requires a TeX distribution, e.g. TinyTeX)

and to different document formats: 

-   word (`format: docx`)
-   pdf (`format: pdf`, N.B. requires a TeX distribution, e.g. TinyTeX)