---
title: "Epidemiological Data Analysis using R - Time-to-event analysis"
author: 
  - Karri Seppä \newline
  - Faculty of Social Sciences, Tampere university \newline
  - Finnish Cancer Registry
institute: "<karri.seppa@cancer.fi>"
format: 
  beamer:
    aspectratio: 32
    navigation: horizontal
    header-includes:  |
      \setbeamertemplate{navigation symbols}{}
      \setbeamertemplate{footline}[page number]{}
editor: visual
---

# Topics somewhat covered

1.  Survival or time to event data & censoring. \medskip
2.  Probability concepts for times to event: \newline  survival, hazard and cumulative hazard functions \medskip
3.  Kaplan--Meier and Nelson--Aalen estimators. \medskip
4.  Regression modelling of hazards: Cox model. \medskip
5.  Competing risks and cumulative incidence function (CIF) \bigskip

Main R functions to be covered (\texttt{survival} package)

-   \texttt{Surv(), survfit(), survminer(), coxph()}

# Survival time -- time to event

Let $T$ be the \textbf{time} spent in a given \textbf{state} from its beginning till a certain \textit{endpoint} or \textit{outcome} \textbf{event} or \textit{transition} occurs, changing the state to another. \newline (\texttt{lex.Cst - lex.dur - lex.Xst})

\bigskip

Examples of such times and outcome events:

-   lifetime: birth $\rightarrow$ death, \medskip

-   duration of marriage: wedding $\to$ divorce, \medskip

-   healthy exposure time: \newline start of exposure $\rightarrow$ onset of disease, \medskip

    \item

    clinical survival time: \newline diagnosis of a disease $\rightarrow$ death.

# Set-up of classical survival analysis

-   \textbf{Two-state model}: only one type of event changes the initial state. \medskip
-   Major applications: analysis of lifetimes since birth and of survival times since diagnosis of a disease until death from any cause.

```{=tex}
\setlength{\unitlength}{0.7pt}
\begin{picture}(400,80)(-40,70)
  \thicklines
  \put(  0, 80){\framebox(110,50){Alive}}
  \put(240, 80){\framebox(110,50){Dead}}
  \put(125,105){\vector(1, 0){100}}
  \put(170,110){\makebox(0,0)[b]{Transition}}
\end{picture}
```
\textbf{Censoring}: Death and final lifetime not observed for some subjects due to emigration or closing the follow-up while they are still alive.

# Distribution concepts: survival function

Cumulative distribution function (CDF) $F(t)$ and density function $f(t) =F'(t)$ of survival time $T$: $$
F(t) =  P( T \le t) = \int_0^t f(u) du 
$$ = \textbf{risk} or probability that the event occurs by $t$.

\bigskip

\textbf{Survival function} $$ 
S(t) = 1- F(t) =  P( T  >  t) = \int_t^{\infty} f(u)du, 
$$ = probability of avoiding the event at least up to $t$ \newline (the event occurs only after $t$).

# Distribution concepts: hazard function

The \textbf{hazard rate} or \textbf{intensity} function $\lambda(t)$ \begin{align*}
\lambda(t) & = \underset{\Delta \rightarrow 0}{\lim} 
 {P(t < T \le t+\Delta | T > t)}/{\Delta} \\
   & = \underset{\Delta \rightarrow 0}{\lim} 
      \frac{P(t < T \le t+\Delta)/\Delta }{P(T > t)}
     = \frac{f(t)}{S(t)}  
\end{align*}

```{=tex}
\begin{enumerate}[$\approx$]
\item  the conditional probability that
the event occurs in a short
 interval $(t, t+\Delta]$, given that it does not
occur before $t$, per interval length. 
\end{enumerate}
```
In other words, during a short interval

```{=tex}
\begin{center}
 risk of event $\approx$ hazard $\times$ interval length 
 \end{center}
```
The \textbf{cumulative hazard} (or integrated intensity): $$
\Lambda(t) = \int_0^t \lambda(v)dv
$$

# Observed data on survival times

For individuals $i = 1, \dots, n$ let \vspace{-0.3cm}

```{=tex}
\begin{enumerate}[ ]
\item $\qquad T_i$ = true time to event,
\item $\qquad U_i$ = true time to censoring.
\end{enumerate}
```
Censoring is assumed \textbf{noninformative}, \textit{i.e.} independent from occurrence of events.

We observe \vspace{-0.3cm}

```{=tex}
\begin{enumerate}[ ]
\item
$y_i = \text{min}\{ T_i, U_i \}$, \textit{i.e.}
the exit time, and
\item
 $ \delta_{i} = 1_{ \{ T_i < U_i  \} }$, 
  indicator (1/0) for the event occurring first, before censoring. 
\end{enumerate}
```
Censoring must properly be taken into account in the statistical analysis.

# Approaches for analysing survival time

```{=tex}
\begin{itemize}
\item 
\textbf{Parametric models} on hazard rate $h(t)$ \\
(like Weibull, gamma, etc.) -- Likelihood:
\begin{align*} L & = \prod_{i=1}^n \lambda(y_i)^{\delta_i} S(y_i) \\
  & = 
   \exp\left\{ \sum_{i=1}^n 
     [ \delta_i \: \log \: \lambda(y_i) - \Lambda(y_i) ] 
       \right\} 
\end{align*}   
\item 
\textbf{Piecewise constant rate} model on $\lambda(t)$ \\ 
-- estimation of $\hat\lambda(t)$ with poisson regression 
\item 
\textbf{Non-parametric} methods, 
like \\ Kaplan--Meier (KM) 
estimator of survival curve $S(t)$ and \newline Cox 
proportional hazards model.
\end{itemize}
```
# R package \texttt{survival}

Tools for analysis with one outcome event.

-   \texttt{Surv(time,event) -> sobj} \newline  creates a \textbf{survival object} \texttt{sobj},

-   \texttt{survfit(sobj) -> sfo} \newline  non-parametric survival curve estimates, like KM (also estimated baseline in a Cox model),

-   \texttt{plot(sfo)} \newline  survival curves and related graphs,

-   \texttt{coxph(sobj \~{} x1 + x2 +...)} \newline   fits the Cox model with covariates \texttt{x1} and \texttt{x2}.

-   \texttt{glm()} -- piecewise constant rate model.

-   \texttt{survreg()} -- parametric survival models.

# KM esimate for survival function $S(t)$

```{=tex}
\begin{itemize}
\item  Order event times (possibly separately in groups)
\item $\widehat{S(t_j)}=\widehat{S(t_{j-1})} \left(1-\frac{d_j}{n_j}\right) $, where $t_0=0$ and
${\widehat{S(0)}=1}$
\item ${\widehat{S(t)}}$ is constant between event -- step function 
\item Each observations contributes as long as at risk for the event and confidence intervals can be introduced using classical inference framework.
\end{itemize}
```
# Veterans' Administration Lung Cancer study

In this trial, males with advanced inoperable lung cancer were randomized to a standard therapy and a test chemotherapy. The primary endpoint for the therapy comparison was time to death in days, represented by the variable Time.

Variables

```{=tex}
\begin{itemize}
\item trt:   1=standard 2=test
\item celltype:1=squamous, 2=smallcell, 3=adeno, 4=large
\item time:  survival time (days)
\item status:    status 1= death, 0=censored 
\item karno:     Karnofsky performance score (100=good)
\item diagtime:months from diagnosis to randomisation
\item age:   in years
\item prior:     prior therapy 0=no, 1=yes
\end{itemize}
```
\scriptsize Reference: D Kalbfleisch and RL Prentice (1980), The Statistical Analysis of Failure Time Data. Wiley, New York.

# Veteran data

\scriptsize

```{r, comment=NA,echo=T}
library(survival)
head(veteran)
head(veteran[veteran$trt==2,])
```

# Estimate for hazard function $\lambda(t)$ in R \newline -- splitting time scale

\scriptsize

```{r, comment=NA,echo=T, warning=F, message=FALSE}
#===============================
# get packages needed for the analysis
#===============================
library(dplyr)
library(survival)
library(Epi)
#===============================
# define stating time at 0 and create subject id
#===============================
veteran$start<-0
veteran$id<-1:nrow(veteran)
#===============================
# Split follow-up time into interval
#===============================
nvet<-survSplit(veteran,
                cut=c(100,200,300,400),
                event="status",
                start="start",
                end="time",episode="period")
```

# Estimate for hazard function $\lambda(t)$ in R \newline -- splitting time scale

\scriptsize

```{r, comment=NA,echo=T, warning=F, message=FALSE}
#===============================
# dplyr -package,rather than stat.table() -function 
#===============================
gsvet<-group_by(nvet,period)
speriod<-summarize(gsvet, 
               n=length(trt),
               events=sum(status),
               pyrs=round(sum(time-start)),
               rate1000=(events/pyrs)*1000,
               lograte=log(events/pyrs))
speriod<-data.frame(speriod)
speriod[,1:6]
```

# Estimate for hazard function $\lambda(t)$ in R \newline -- Poisson regression

\scriptsize

```{r, comment=NA,echo=T, warning=F, message=FALSE}
#===============================
# Poisson model for incidence in each period of time
#===============================
m<-glm(events ~ -1+factor(period),
       family=poisson,offset=log(pyrs),data=speriod)
round(ci.lin(m,Exp=TRUE)[,-(3:4)],6)
#1-year cumulative hazard 
Lambda_1yr<- sum(exp(m$coefficient[1:4])*c(100,100,100,65))
#1-year survival in percentages
exp(-Lambda_1yr)*100
```

# Estimation of the survival function $S(t)$

\scriptsize

```{r, comment=NA,echo=T, warning=F, message=FALSE}
m <- survfit(formula = Surv(time, status) ~ trt, data = veteran)
m2 <- summary(m)
m2 <- data.frame(lapply(c(10,2:6) , function(x) m2[x]))
m2[m2$time<10,]
```

::: columns
::: {.column width="50%"}
-   trt=1

```{=tex}
\begin{enumerate}[ ]
\item $S(0)=1$\\
\item $S(3)=S(0)\times(1-1/69)=0.986$\\
\item $S(4)=S(3)\times(1-1/68)=0.971$
    
\end{enumerate}
```
:::

::: {.column width="50%"}
-   trt=2

```{=tex}
\begin{enumerate}[ ]
\item $S(0)=1$\\
\item $S(1)=S(0)\times(1-2/68)=0.971$\\
\item $S(2)=S(1)\times(1-1/66)=0.956$

\end{enumerate}
```
:::
:::

# Plot of Survival curve for Veteran data

\scriptsize

```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=6.5}
library(survminer)
ggsurvplot(m, pval = TRUE, conf.int = TRUE, risk.table = TRUE)
```

# Regression models for time-to-event data: Cox model

Baseline hazard $\lambda_0(t,\alpha)$ may be given a parametric form (\textit{e.g.} Weibull) or a piecewise constant rate (exponential) structure.

\bigskip

Often a parameter-free form $\lambda_0(t)$ is assumed. Then $$
  \lambda_i(t) = \lambda_0(t) \exp(\eta_1),
$$ specifies the \textbf{Cox model} or the \textbf{semiparametric proportional hazards model}.

\bigskip

$\eta_i = \beta_1 x_{i1} + \dots + \beta_p x_{ip}$ not depending on time.

\bigskip

Generalizations: \textbf{time-dependent} \newline covariates $x_{ij}(t)$, and/or effects $\beta_j(t)$.

# PH model: interpretation of parameters

Present the model explicitly in terms of $x$'s and $\beta$'s. $$
\lambda_i(t) = \lambda_0(t)  \exp({\beta_1 x_{i1} + \dots +
\beta_p x_{ip}})
$$ Consider two individuals, $i$ and $i'$, having the same values of all other covariates except the $j^{\text{th}}$ one.

\bigskip

The ratio of hazards is constant: $$  \frac{\lambda_i(t)}{\lambda_{i'}(t)} = \frac{\exp( \eta_{i}) }{\exp(\eta_{i'})}
= \exp \{ \beta_j(x_{ij}-x_{i'j}) \} . $$ Thus $e^{\beta_j} = \text{HR}_j$ = \textbf{hazard ratio} or relative rate associated with a unit change in covariate $X_j$.

# Ex. Veteran data and treatment effect

Fitting Cox models with trt effect. \scriptsize

```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4.5}
m <- coxph(formula = Surv(time, status) ~ trt, data = veteran)
summary(m)
```

\normalsize

HR for treatment 2 vs. 1 is 1.02 (95% CI 0.71;1.45) \newline Not statistically significant (p=0.92)

# Proportionalilty of hazards?

```{=tex}
\begin{itemize}
\item
Consider two groups $g$ and $h$ defined by one categorical covariate, and let $\rho > 0$. 
\medskip
If $\lambda_g(t) = \rho \lambda_h(t)$  then 
$\Lambda_g (t) = \rho \Lambda_h(t)$ and
$$ \log\: \Lambda_g  (t)  = \log (\rho) + \log\: \Lambda_h (t), $$
thus log-cumulative hazards should be parallel!
\end{itemize}
\begin{enumerate}[$\Rightarrow$]
\item
\textit{Plot the estimated log-cumulative hazards and see
whether they are sufficiently parallel}.
\end{enumerate}
\begin{itemize}
\item
\texttt{plot(coxobj, ..., fun = 'cloglog')}
\medskip
\item
Testing the proportionality assumptions: \texttt{cox.zph(coxobj)}.
\end{itemize}
```
# Ex. Veteran data - test PH

```{=tex}
\begin{itemize}
\item With $>1$ covariates, \texttt{cox.zph()} tests the assumption by checking, whether the corresponding parameters (\& hazard ratios) may vary in time.
\item Suppose that I want to include information on patient baseline general disease status -- Karnofsky performance score (0=dead, 100=good)
\item Dichotomize the Karnofsky score 0 if Score $[$0,50 $]$ and 1 if $($50,100$]$
\end{itemize}
```
# Ex. Veteran data - test PH

\scriptsize

```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4.5}
veteran$karnod<-as.numeric(veteran$karno>50)
m <- coxph(formula = Surv(time, status) ~ trt+ karnod, 
           data = veteran)
m
#testing proportionality
cox.zph(m)
```

Test for proportionality are significant (p\<0.05) -- assumptions of proportionality of hazards is rejected for the score variable -- stratify baseline hazard by the score

# Ex. Veteran data - test PH

\scriptsize

```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4.5}
m <- coxph(formula = Surv(time, status) ~ trt+strata(karnod), 
           data = veteran)
m
cox.zph(m)
```

Test for proportionality is not significant (p\>0.05) \newline HR for treatment effect is 1.01 95% CI (0.71 ;1.45)

# Competing events & competing risks

In many epidemiological and clinical contexts there are competing events that may occur before the target event and remove the person from the population at risk for the event, \textit{e.g.}

```{=tex}
\begin{itemize}
\item \textit{target event}: occurrence of endometrial cancer,
 \textit{competing events}: hysterectomy or death.
\medskip
\item \textit{target event}: relapse of a disease \\
(ending the state of remission), \\
 \textit{competing event}: death while still in remission.
 
\item \textit{target event}: divorce, \\ 
 \textit{competing event}: death of either spouse.
 
\end{itemize}
```
# Competing risks model: causes of death

```{=tex}
\begin{itemize}
\item
Often the interest is focused on the risk or hazard of dying 
from one specific cause.
\medskip
\item
That cause may eventually not be realized, because
a \textbf{competing cause} of death hits first.

\bigskip

\begin{center}
\setlength{\unitlength}{0.65pt}
\begin{picture}(400,190)
  \thicklines
  \put(  0, 80){\framebox(120,50){Alive}}
  \put(230,130){\framebox(180,50){Dead from cancer}}
  \put(230, 30){\framebox(180,50){Dead, other causes}}
  \put(135,125){\vector(3, 1){90}}
  \put(135, 85){\vector(3,-1){90}}
  \put(165,157){\makebox(0,0)[b]{$\lambda_1(t)$}}
  \put(165, 53){\makebox(0,0)[t]{$\lambda_2(t)$}}
\end{picture}
\end{center}
\item
 Generalizes to several competing causes.
 \end{itemize}
```
# Event-specific quantities

\textbf{Cumulative incidence function} (CIF) for event $c$: $$  F_c(t) = P(T \leq t \text{ and } C = c ), \quad c = 1,2.  $$

\medskip

From this one can recover

```{=tex}
\begin{itemize}
\item
$F(t) = \sum_{c} F_c(t)$, CDF of event-free survival time $T$, \textit{i.e.} 
cumulative risk of any event by $t$.
\medskip
\item
$S(t) = 1 - F(t)$, \textbf{event-free survival function}, \textit{i.e.} probability of avoiding all events by $t$
\end{itemize}
```
# Event-specific quantities (cont'd)

\textbf{Event-} or \textbf{cause-specific hazard function} \begin{align*}
 \lambda_c(t) & =  \underset{\Delta\to 0}{\lim} 
    \frac{P(t < T \le t+\Delta \text{ and } C = c \mid T > t)}{\Delta}  \\
        & =  \frac{f_c(t)}{1-F(t)}
\end{align*}

```{=tex}
\begin{enumerate}[$\approx$]
\item Risk of \textit{event} $c$ 
in a short interval $(t, t+ \Delta]$,  given 
\textit{avoidance of all events} up to $t$, 
per interval length.
\end{enumerate}
```
```{=tex}
\begin{itemize}
\item CIF = risk of event $c$ over risk period $[0,t]$ in the presence of competing risks, also obtained 
$$ F_c(t) = \int_0^t \lambda_c(v) S(v) dv, \quad c = 1,2, $$
\item
Depends on the hazard of the competing event, too, 
via 
\begin{align*} 
S(t) & = \exp\left\{ - \int_0^t [\lambda_1(v) + \lambda_2(v)]dv 
 \right\} .
\end{align*}
\end{itemize}
```
# Non-parametric estimation of CIF

```{=tex}
\begin{itemize}
\item
Let $t_1 < t_2 < \dots < t_K$ be the $K$ distinct 
time points at which any outcome event was observed, \\ Let also
 $\widetilde{S}(t)$ be KM estimator for overall $S(t)$. 
\medskip
\item
\textbf{Aalen-Johansen estimator} (AJ) for the cumulative incidence function $F(t)$
is obtained as
$$ 
\widetilde{F}_c(t) 
   = \sum_{t_k \leq t} \frac{D_{kc}}{n_k} \times \widetilde{S}(t_{k-1}),
   \quad\text{where}
$$
$n_k$ = size of the risk set at $t_k$ $(k=1, \dots, K)$,
\\
$D_{kc}$ = no. of cases of event $c$ observed  at $t_k$.
\medskip
\item
\textbf{N.B.} Naive KM estimator of ``cause-specific survival'' treats  
competing events occuring first as censorings.
\end{itemize}
```
# Ex. Survival of 338 oral cancer patients

Important variables:

```{=tex}
\begin{itemize}
\item \texttt{time} = duration of patientship from \\ 
 diagnosis (\textbf{entry}) till death or censoring, 
\medskip
\item
\texttt{event} = indicator for the outcome and its \\
 observation at the end of follow-up (\textbf{exit}): \\
  0 = censoring,  \\
  1 = death from oral cancer, \\
  2 = death from some other cause.
\end{itemize}
```
\medskip

Special features:

```{=tex}
\begin{itemize}
\item
   Several possible endpoints, \textit{i.e.} alternative causes of death, of which
   only one is realized.
   \medskip
\item
   Censoring -- incomplete observation of the survival time.   
\end{itemize}
```
# Ex. Oral cancer data

```{r, comment=NA,echo=T}
orca <- read.table(file="oralca2.txt")
head(orca)
```

# Ex. Survival from oral cancer

```{=tex}
\begin{itemize}
\item
Creating a {\tt Lexis} object with two outcome events and \\ 
obtaining a summary of transitions
\end{itemize}
\scriptsize
```
```{r, comment=NA,echo=T,message=FALSE,warning=FALSE,results='hide'}
library(Epi)
orca.lex <- Lexis(exit = list(stime = time), 
           exit.status = factor(event, 
    labels = c("Alive", "Oral ca. death", "Other death") ),
                  data = orca)
summary(orca.lex)  
```

```{r, comment=NA,echo=F,message=FALSE,warning=FALSE}
summary(orca.lex)  
```

# Box diagram for transitions

Interactive use of function \tt boxes().
\small
```{r, comment=NA,echo=T,message=FALSE,warning=FALSE,eval=FALSE}
boxes(orca.lex)
```

```{=tex}
\begin{center}
\includegraphics[height=6cm]{orca-boxes}
\end{center}
\normalsize
```
# Ex. Survival from oral cancer

```{=tex}
\begin{itemize}
\item
AJ-estimates of CIFs (solid) for both causes.
\item
Naive KM-estimates of CIF (dashed) $>$ AJ-estimates 
\item
CIF curves may also be stacked (right).  
\end{itemize}
```
\includegraphics[width=10cm]{orcaCI1}

\textbf{NB.} The sum of the naive KM-estimates of CIF exceeds 100% at 13 years!

# Warning of "net risk" and "cause-specific survival"

```{=tex}
\begin{itemize}                              
\item
The ``\textbf{\textit{net risk}}'' of outcome $c$ by time $t$, 
assuming hypothetical elimination of competing risks,
is often defined as
\begin{center}
$ F_c^*(t) = 1 - S_c^*(t) = 1- \exp\left\{ - \int_0^t [\lambda_c(v)]dv \right\} $
\end{center}
\medskip
\item
In clinical survival studies, function 
$S_c^*(t)$ is often called ``\textbf{\textit{cause-specific survival}}'',
and estimated by KM, but treating competing deaths
as censorings.
\medskip
\item
Yet, these *-functions, $ F_c^*(t)$ and $S_c^*(t)$, lack proper probability interpretation when
 competing risks exist.
\medskip
\item
Hence, their use and naive KM estimation should be viewed critically
(Andersen \& Keiding, \textit{Stat Med}, 2012)
\end{itemize}
```
